---
title: "R Notebook"
output: html_notebook
---
## Data loading
```{r}
## Data import
data<-read.csv("/Users/yiraozhang/Desktop/cyanobacteria_pgm/cyanobacterial-bloom-prediction/cyano&spw.csv")
data$Cyano<-as.double(data$Cyano)
vars <- c("DO", "ALLSKY_SFC_SW_DWN", "PRECTOT","WS10M_RANGE","pH","elev","Temp","TN","TP","Chl.a","TNTP","NONH","Turb","Fe","TKN","Spc","Cyano")
data <- data[vars]
colnames(data)<-c("DO", "Solar_radiation", "Precipitation","Wind","pH","Elevation","Temp","TN","TP","Chl.a","TNTP","NONH","Turb","Fe","TKN","Spc","Cyano")
data
```
## Data preprocessing
https://www.waterra.com.au/cyanobacteria-manual/Chapter6.htm
Threshold definitions for a general Alert Levels Framework for management of toxic cyanobacteria in drinking water:
**Low alert**: <br>
Detection> 500 & < 2,000 cells mL^-1^
**Medium alert**:<br>
Detection >2,000(2) & < 6,500 cells mL^-1^
**High alert**:<br>
Detection > 6,500 cells mL^-1^
**Very high alert**<br>
Detection > 65,000 cells mL^-1^
```{r}
## Categorize
data$category<-"safe"
data$category[data$Cyano>500 & data$Cyano<2000] <- "low"
data$category[data$Cyano>2000 &data$Cyano<6500] <- "medium"
data$category[data$Cyano>6500 &data$Cyano<65000] <- "high"
data$category[data$Cyano>65000] <- "veryhigh"
## Split training/test set
library(caTools)
set.seed(100) 
sample = sample.split(data, SplitRatio = .8)
train = subset(data, sample == TRUE)
test  = subset(data, sample == FALSE)
data<-train
```
## Data discription and visiualization
```{r}
library(ggplot2)
library(ggthemes)
library(cowplot)
## Attribute Histograms
tema = theme(plot.background = element_rect(fill="#E0FFFF"),
             axis.title.x = element_text(size=8, color = "black"),
             axis.title.y = element_text(size=8, color = "black"),
             axis.text.x = element_text(size=5),
             axis.text.y = element_text(size=5))
do<-ggplot(data = data, mapping = aes(x = DO)) +
    geom_histogram(bins=30, fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("DO") +
    tema
wind<-ggplot(data = data, mapping = aes(x = Wind)) +
    geom_histogram(bins=30, fill = "blue", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Wind") +
    tema
solar<-ggplot(data = data, mapping = aes(x = Solar_radiation)) +
    geom_histogram(bins=30, fill = "green", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Solar") +
    tema
prec<-ggplot(data = data, mapping = aes(x = Precipitation)) +
    geom_histogram(bins=30, fill = "yellow", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Precipitation") +
    tema
ph<-ggplot(data = data, mapping = aes(x = pH)) +
    geom_histogram(bins=30, fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("pH") +
    tema
elev<-ggplot(data = data, mapping = aes(x = Elevation)) +
    geom_histogram(bins=30, fill = "blue", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Elevation") +
    tema
temp<-ggplot(data = data, mapping = aes(x = Temp)) +
    geom_histogram(bins=30, fill = "green", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Temperature") +
    tema
tn<-ggplot(data = data, mapping = aes(x = TN)) +
    geom_histogram(bins=30, fill = "yellow", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Total Nitrogen") +
    tema
tp<-ggplot(data = data, mapping = aes(x = TP)) +
    geom_histogram(bins=30, fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("TP") +
    tema
tntp<-ggplot(data = data, mapping = aes(x = TNTP)) +
    geom_histogram(bins=30, fill = "blue", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("TNTP") +
    tema
chl.a<-ggplot(data = data, mapping = aes(x = Chl.a)) +
    geom_histogram(bins=30, fill = "green", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Chl.a") +
    tema
nonh<-ggplot(data = data, mapping = aes(x = NONH)) +
    geom_histogram(bins=30, fill = "yellow", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("NONH") +
    tema
turb<-ggplot(data = data, mapping = aes(x = Turb)) +
    geom_histogram(bins=30, fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Turbidity") +
    tema
fe<-ggplot(data = data, mapping = aes(x = Fe)) +
    geom_histogram(bins=30, fill = "blue", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Fe") +
    tema
tkn<-ggplot(data = data, mapping = aes(x = TKN)) +
    geom_histogram(bins=30, fill = "green", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Total Kjeldahl Nitrogen") +
    tema
spc<-ggplot(data = data, mapping = aes(x = Spc)) +
    geom_histogram(bins=30, fill = "yellow", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Specific Conductance") +
    tema
cyano<-ggplot(data = data, mapping = aes(x = Cyano)) +
    geom_histogram(bins=30, fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Cyanobacteria Abundance") +
    tema
plot_grid(do, wind, solar, prec, ph, elev, temp, tn, tp, tntp, chl.a, nonh, turb, fe, tkn, spc, nrow=4, ncol=4)
plot_grid(cyano)
```
```{r}
## Density plot and boxplot
tema = theme(plot.background = element_rect(fill="#E0FFFF"),
             axis.title.x = element_text(size=8, color = "black"),
             axis.title.y = element_text(size=8, color = "black"),
             axis.text.x = element_text(size=5),
             axis.text.y = element_text(size=5))
do<-ggplot(data = data, mapping = aes(x = DO)) +
    geom_density( fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("DO") +
    tema
wind<-ggplot(data = data, mapping = aes(x = Wind)) +
    geom_density( fill = "blue", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("Wind") +
    tema
solar<-ggplot(data = data, mapping = aes(x = Solar_radiation)) +
    geom_density( fill = "green", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("Solar") +
    tema
prec<-ggplot(data = data, mapping = aes(x = Precipitation)) +
    geom_density( fill = "yellow", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("Precipitation") +
    tema
ph<-ggplot(data = data, mapping = aes(x = pH)) +
    geom_density( fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("pH") +
    tema
elev<-ggplot(data = data, mapping = aes(x = Elevation)) +
    geom_density( fill = "blue", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("Elevation") +
    tema
temp<-ggplot(data = data, mapping = aes(x = Temp)) +
    geom_density( fill = "green", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("Temperature") +
    tema
tn<-ggplot(data = data, mapping = aes(x = TN)) +
    geom_density( fill = "yellow", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("Total Nitrogen") +
    tema
tp<-ggplot(data = data, mapping = aes(x = TP)) +
    geom_density( fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("TP") +
    tema
tntp<-ggplot(data = data, mapping = aes(x = TNTP)) +
    geom_density( fill = "blue", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("TNTP") +
    tema
chl.a<-ggplot(data = data, mapping = aes(x = Chl.a)) +
    geom_density( fill = "green", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("Chl.a") +
    tema
nonh<-ggplot(data = data, mapping = aes(x = NONH)) +
    geom_density( fill = "yellow", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("NONH") +
    tema
turb<-ggplot(data = data, mapping = aes(x = Turb)) +
    geom_density( fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("Turbidity") +
    tema
fe<-ggplot(data = data, mapping = aes(x = Fe)) +
    geom_density( fill = "blue", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("Fe") +
    tema
tkn<-ggplot(data = data, mapping = aes(x = TKN)) +
    geom_density( fill = "green", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("Total Kjeldahl Nitrogen") +
    tema
spc<-ggplot(data = data, mapping = aes(x = Spc)) +
    geom_density( fill = "yellow", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("Specific Conductance") +
    tema
cyano<-ggplot(data = data, mapping = aes(x = Cyano)) +
    geom_density( fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist()+
    xlab("Cyanobacteria Abundance") +
    tema
plot_grid(do, wind, solar, prec, ph, elev, temp, tn, tp, tntp, chl.a, nonh, turb, fe, tkn, spc, nrow=4, ncol=4)
plot_grid(cyano)
```
```{r}
## Correlation graph
library(GGally)
df <- subset(data, select = -c(category, Cyano))
colnames(df)<-c("DO", "Solar", "Precipitation","Wind","pH","Elevation","Temp","TN","TP","Chl.a","TNTP","NONH","Turb","Fe","TKN","Spc")
ggcorr(df)
```
## Log-transform variables
```{r}
data_transformed<-data
data_transformed$DO<-log(data_transformed$DO)
data_transformed$Solar_radiation<-log(data_transformed$Solar_radiation)
data_transformed$TN<-log(data_transformed$TN)
data_transformed$TP<-log(data_transformed$TP)
data_transformed$Chl.a<-log(data_transformed$Chl.a)
data_transformed$pH<-log(data_transformed$pH)
data_transformed$TNTP<-log(data_transformed$TNTP)
data_transformed$Temp<-log(data_transformed$Temp)
data_transformed$Cyano<-log(data_transformed$Cyano)
data_transformed$Elevation<-log(data_transformed$Elevation)
data_transformed$TKN<-log(data_transformed$TKN)
data_transformed$Fe<-log(data_transformed$Fe)
data_transformed$NONH<-log(data_transformed$NONH)
data_transformed$Turb<-log(data_transformed$Turb)
data_transformed$Spc<-log(data_transformed$Spc)
data_transformed$Wind<-log(data_transformed$Wind)
tema = theme(plot.background = element_rect(fill="#E0FFFF"),
             axis.title.x = element_text(size=8, color = "black"),
             axis.text.x = element_text(size=5),
             axis.text.y = element_text(size=5))
do<-ggplot(data = data_transformed, mapping = aes(x = DO)) +
    geom_histogram(bins=30, fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("DO") +
    tema
wind<-ggplot(data = data_transformed, mapping = aes(x = Wind)) +
    geom_histogram(bins=30, fill = "blue", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Wind") +
    tema
solar<-ggplot(data = data_transformed, mapping = aes(x = Solar_radiation)) +
    geom_histogram(bins=30, fill = "green", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Solar") +
    tema
prec<-ggplot(data = data_transformed, mapping = aes(x = Precipitation)) +
    geom_histogram(bins=30, fill = "yellow", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Precipitation") +
    tema
ph<-ggplot(data = data_transformed, mapping = aes(x = pH)) +
    geom_histogram(bins=30, fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("pH") +
    tema
elev<-ggplot(data = data_transformed, mapping = aes(x = Elevation)) +
    geom_histogram(bins=30, fill = "blue", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Elevation") +
    tema
temp<-ggplot(data = data_transformed, mapping = aes(x = Temp)) +
    geom_histogram(bins=30, fill = "green", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Temperature") +
    tema
tn<-ggplot(data = data_transformed, mapping = aes(x = TN)) +
    geom_histogram(bins=30, fill = "yellow", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Total Nitrogen") +
    tema
tp<-ggplot(data = data_transformed, mapping = aes(x = TP)) +
    geom_histogram(bins=30, fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("TP") +
    tema
tntp<-ggplot(data = data_transformed, mapping = aes(x = TNTP)) +
    geom_histogram(bins=30, fill = "blue", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("TNTP") +
    tema
chl.a<-ggplot(data = data_transformed, mapping = aes(x = Chl.a)) +
    geom_histogram(bins=30, fill = "green", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Chl.a") +
    tema
nonh<-ggplot(data = data_transformed, mapping = aes(x = NONH)) +
    geom_histogram(bins=30, fill = "yellow", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("NONH") +
    tema
turb<-ggplot(data = data_transformed, mapping = aes(x = Turb)) +
    geom_histogram(bins=30, fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Turbidity") +
    tema
fe<-ggplot(data = data_transformed, mapping = aes(x = Fe)) +
    geom_histogram(bins=30, fill = "blue", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Fe") +
    tema
tkn<-ggplot(data = data_transformed, mapping = aes(x = TKN)) +
    geom_histogram(bins=30, fill = "green", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Total Kjeldahl Nitrogen") +
    tema
spc<-ggplot(data = data_transformed, mapping = aes(x = Spc)) +
    geom_histogram(bins=30, fill = "yellow", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Specific Conductance") +
    tema
cyano<-ggplot(data = data_transformed, mapping = aes(x = Cyano)) +
    geom_histogram(bins=30, fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Cyanobacteria Abundance") +
    tema
plot_grid(do, wind, solar, prec, ph, elev, temp, tn, tp, tntp, chl.a, nonh, turb, fe, tkn, spc, nrow=4, ncol=4)
plot_grid(cyano)
```
## Subset with Cyanobacteria abundance>0
```{r}
df<-data_transformed[data_transformed$Cyano>0,]
cyano<-ggplot(data = df, mapping = aes(x = Cyano)) +
    geom_histogram(bins=30, fill = "red", color = "black", size = 1.0, alpha = .5) +
    theme_economist() +
    xlab("Cyanobacteria Abundance") +tema
plot_grid(cyano)
```
## Normality Visual Check with Q-Q plot
```{r}
library(car)
par(mfrow = c(2, 3))
qqPlot(data_transformed$pH)
qqPlot(data_transformed$DO)
qqPlot(data_transformed$Wind)
qqPlot(data_transformed$TN)
qqPlot(data_transformed$TP)
qqPlot(data_transformed$TNTP)
par(mfrow = c(2, 3))
qqPlot(data_transformed$NONH)
qqPlot(data_transformed$Turb)
qqPlot(data_transformed$Chl.a)
qqPlot(data_transformed$Cyano)
```
## ## Structure learning with **bnlearn**
```{r}
library(bnlearn)
library(Rgraphviz)
df <- subset(df, select = -c(Fe, TKN, category, Spc))
bl=matrix(c("Turb","Temp","Turb","DO","DO","Temp","Chl.a","Temp","Elevation","pH","pH","Elevation","Elevation","TN","TN","Elevation","pH","Temp","Solar_radiation","Elevation","Cyano","Temp","Cyano","Elevation","Elevation","Cyano", "Chl.a", "Cyano","Solar_radiation","Turb", "Precipitation","Temp","Chl.a","pH"), ncol = 2, byrow = TRUE)
net1<-tabu(df, blacklist = bl)
graphviz.plot(net1)
```
## Generalized Linear Model
```{r}
data$Cyano[data$Cyano==1] <- 0
test$Cyano[test$Cyano==1] <- 0
# zero-inflated data
length(which(data$Cyano==0))/length(data$Cyano)
## GLM Poisson
library(MASS)
model_poisson <- glm(Cyano ~ TP + pH + Temp + Solar_radiation + Chl.a,
          family = 'poisson',
          data = data)
res <- resid(model_poisson, type = "pearson")
N  <- nrow(data)
p  <- length(coef(model_poisson))   
sum(res^2) / (N - p)
## GLM Negative Binomial
model_negbin <- glm.nb(Cyano ~ TP + pH + Temp + Solar_radiation + Chl.a,
          data = data)
## Zero-inflated NB
library(pscl)
model_ZINB <- zeroinfl(Cyano ~ TP + pH + Temp + Solar_radiation + Chl.a|
                 TP + pH + Temp + Solar_radiation + Chl.a,
               dist = 'negbin',
               data = data)
## Hurdle NB
model_hurdleNB <-hurdle(Cyano ~ TP + pH + Temp + Solar_radiation + Chl.a|
                 TP + pH + Temp + Solar_radiation + Chl.a,
               dist = 'negbin',
               data = data)
```
## Model selection
```{r}
AIC(model_negbin, model_ZINB, model_hurdleNB)
```

```{r}
library(pscl)
vuong(model_negbin, model_hurdleNB)
vuong(model_negbin, model_ZINB)
vuong(model_ZINB, model_hurdleNB)
```

## MCMC
```{r}
library(dplyr)
library(magrittr)
library(rstan)
x<-data %>% select(TP, Solar_radiation, pH, Temp, Chl.a)
x_test<-test%>%select(TP, Solar_radiation, pH, Temp, Chl.a)
stan_data <- list(N=nrow(x),
                  K=ncol(x),
                  x=x,
                  y=data$Cyano,
                  N_test=nrow(x_test),
                  x_test=x_test)
fit <- stan(file = 'ZINB.stan', data = stan_data, iter = 1000, chains = 4)
```
```{r}
head(summary(fit)$summary, n=13L)
```
```{r}
plot(fit)
pairs(fit, pars = c("intercept1", "intercept2", "phi"))
pairs(fit, pars = c("beta_mu[2]","beta_mu[4]"))
pairs(fit, pars = c("beta_theta[3]","beta_theta[5]"))
```
## Colinearity
```{r}
cor(x$pH, x$Chl.a)
cor(x$Solar_radiation, x$Temp)
```
```{r}
traceplot(fit, inc_warmup = TRUE)
```
## Posterior Predictive Checks (PPCs)
The idea behind posterior predictive checking is simple: if a model is a good fit then we should be able to use it to generate data that looks a lot like the data we observed. 
```{r}
library(bayesplot)
y_rep <- as.matrix(fit, pars = "y_pre")
color_scheme_set("brightblue")
ppc_dens_overlay(test$Cyano, y_rep[1:200, ])+ xlim(0, 100000)
#Here we see data (dark blue) fit well with our posterior predictions.
```
```{r}
ppc_stat(y = test$Cyano, yrep = y_rep, stat = "mean") + xlim(0, 100000)
```
```{r}
prop_zero <- function(x) mean(x == 0)
ppc_stat(y = test$Cyano, yrep = y_rep, stat = "prop_zero") + xlim(0, 1)
```
## Evalute and compare the regression models
```{r}
library(berryFunctions)
library(Metrics)
#zero-inflated negative binomial glm model
predict1 <-as.data.frame(predict(model_ZINB,newdata=test,type='response'))
colnames(predict1)<-c("Cyano")
rsq1<-rsquare(predict1$Cyano, test$Cyano)
rmse1<-rmse(predict1$Cyano, test$Cyano)
mae1<-mae(predict1$Cyano, test$Cyano)
#Bayesian zero-inflated negative binomial glm model
predict2<-as.data.frame(apply(y_rep, 2, median))
colnames(predict2)<-c("Cyano")
rsq2<-rsquare(predict2$Cyano, test$Cyano)
rmse2<-rmse(predict2$Cyano, test$Cyano)
mae2<-mae(predict2$Cyano, test$Cyano)
#result table
result<-matrix(c(rsq1, rmse1, mae1, rsq2, rmse2, mae2), ncol=3, byrow=TRUE)
colnames(result) <- c("R_squared","RMSE","MAE")
rownames(result) <- c("ZINB","Bayesian ZINB")
result <- as.table(result)
result
```
#Sensitivity Analysis
```{r}
```